\chapter{Introduction}
Least trimmed squares (LTS) is one of many modifications very well known method ordinary least squares (OLS). Both of these methods are tools of regression analysis, which is a group of processes used to estimate the dependence of variables. Specifically in the case when we try to estimate dependence one variable on one or multiple others. The regression analysis uses the regression models, and in the case of OLS and LTS methods, that model is the linear regression model.

To OLS estimate produces reliable results, many strong assumptions about the data have to be fulfilled. It is assumed that data is generated specific way as well as that the data does not contain measurement errors called outliers. Those assumptions are in practice hardly fulfilled, because outliers in the data are very common. The OLS estimate is in such cases unreliable.

Problems of classical statistic methods try to solve robust statistic. Its methods are usually emulations of classical statistic methods but try to provide good performance even if data contains a large number of outliers. That means such methods does not rely so much on assumptions which are difficult to achieve in practice. Because the OLS method is one of the essential tools of linear regression analysis, multiple its alternatives have been designed to fulfill the assumptions of a robust estimator.

One of the methods is LTS. Its idea is simple, but unlike the OLS method, the exact formula for calculation is not known, so that only to date known exact algorithms for computing the LTS estimate are combinatorial. For this reason beside exact algorithms, several probabilistic algorithms have been proposed.

This work is divided into three chapters. In the first chapter, we introduce theory required to understand linear regression model, OLS and LTS. We mention properties of both methods and also describe filed of its usage.

In the second chapter, we cover algorithms for calculating the OLS estimate. It is necessary because most of the algorithm used to compute the LTS estimate relies on those algorithms. Next, we describe all currently used algorithms for calculating LTS estimate. They consist of several probabilistic and also few exact ones. In this chapter, we also show that those algorithms can be easily combined to obtain higher speed and performance.

In the last chapter, we describe our experimental results. At first, we cover data generator which is used for our experiments and which can provide data affected by various types of outliers. Next, we provide information about our implementation of all algorithms for chapter two. Finally, we present our results for specific data.